---
title: "R for Successful Student Projects 02 - Tabular Data Basics"
date: "Semester 1, 2022"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)
```

 
## Tidy data principles

Before we delve we bring data into R and start to visualise it. We first need to learn about how computers like data to be structured.

It's very common to collect data using a spreadsheet, so here are some principles for setting up your spreadsheet:

1. Make it rectangular
2. Columns are variables
3. Rows are observations
4. One piece of information per cell

Think about how you are dealing with missing data.

In general we would recommend saving your data in the "comma separated values" (csv) format. This is a format where all your data is in plain text (you can view it in a text editor) and it is highly portable between different languages/programs.


## Read data in from csv

read.csv and read_csv

- stringsAsFactors=FALSE
- prechecks
  - `dim`
  - `View`
  - `head`
  - `tail`
  - `names`


## Data visualisation {.tabset}

There are two main methods for plotting data in R. The first is the in-built plotting functions within base R. The second is using a popular package called `ggplot2`.

Here we'll introduce you to both systems to create some simple plots. In future modules we'll cover more about plotting.

### Base R 

#### Scatter plots
```{r}
plot(mtcars$mpg, mtcars$wt)
```

#### Line plots

```{r}
plot(mtcars$hp, mtcars$mpg)
```


#### Histograms

```{r}
hist(mtcars$mpg)
```


### Tidyverse

The ggplot2 package is a popular data visualisation package for R. In order to use it we need to first load it into R.

```{r}
library(ggplot2)
```



#### Creating a plot

```{r}
ggplot(data = mtcars, aes(x = mpg, y = wt)) + geom_point()
```


- scatter plot
- line plot


