---
title: "Get Started with R"
author: "Patricia Haden & Murray Cadzow"
date: "Semester 1, 2022"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)


```

\

## R and RStudio

R is the language, RStudio is an interface for interacting with R

## The RStudio environment

Panels of RStudio

- console
- source
- environment/history
- file/plot/help/viewer

### Projects



### Scripts

R Script

RMarkdown

## R syntax

General syntax info:

- R is case sensitive
- spaces are ignored by R
- there is no difference between single `'` and double `"` quotes but you have to use the same type to close as you open with


### Mathematical operators

addition: `+`

subtraction: `-`

multiplication: `*`

division: `/`

exponent: `**`

modulo/remainder: `%%`

```{r}
# addition
1 + 2

# subtraction
3 - 6

# multiplication
4 * 2

# division
12 / 3

# exponent
2 ** 5

# modulo
5 %% 2

```


### Variables

- `<-` and `=`
- last thing assigned overwrites

### Functions

- arguments
- `args()`

### Help

- `?` and `??`
- `help()`
- google

### Vectors

vectors

- use of [] for subsetting
- subsetting by index
- atomic data types
  - character
  - double
  - logical
  - `typeof()`


### Comparitors

- logic statements
- comparitor subsetting

- equality: `==`
- not equal to:`!=`
- greater than / greater than or equal to: `>` / `>=`
- less than / less than or equal to: `<` / `<=`


- not: `!`
- or: `|`
- and: `&`

