---
title: "Handout 03 - Selecting and Filtering Data"
author: "Patricia Haden & Murray Cadzow"
date: "Semester 1, 2022"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)
```


## Vector subsetting

In the [first week](handout1.html) we touched on vectors and using the `[]` to extract items out based on an index. We're now going to expand on this idea but for our dataframes.

Often it's fairly laborious to work out exactly which index corresponds to the data you're interested in. Usually we're more interested in data that meets certain _conditions_. Instead of sub-setting by indices, we can do conditional sub-setting.


Lets create an example vector to work use
```{r}
nums <- c(1, 2, 3, 4, 5, 6)

nums
```

Similar to sub-setting by index, we can provide a vector of booleans (`TRUE`/`FALSE`) to `[]` to specify which elements to keep.

```{r}
nums[c(TRUE, FALSE, TRUE, FALSE, TRUE, FALSE)]
```



We can apply mathematical operators and our comparative operators to entire vectors. When using a comparator we get back a vector of equal length of booleans.

```{r}
nums < 4
```

We can use this vector to filter our vectors to keep only the values that returned `TRUE`.

```{r}
nums[nums < 4]
```


## Dataframe subsetting {.tabset}

### Base R

- `$`
- `[,]`
  - names
  - index
  - conditional




### dplyr

pipes %>% 

- select
  - starts_with
  - ends_with
  - contains
  - anyof


- filter
- slice





## Plotting

base R and ggplot2

- 
- line plot
