---
title: 'Exercises R4SSP Module 1: Get Started with R '
date: "Semester 1, 2022"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)


```

# Exercises
Write R code to solve the following exercises. Save all your code in an R script file. Our solutions are at the end of the handout if you want to compare.

If you have any comments, questions or problems, email us, or come to the scheduled session on Monday between 10.00 and 12.00 in the Science Library Seminar Room.

1. Your flatmate is doing study exchange in Arkansas (in the United States). She messages you that it is -17 degrees Fahrenheit. To convert a Fahrenheit temperature reading to Celsius, you subtract 32 from it and then multiply that value by 5/9. How cold is your flatmate in Celsius? Hint: You can group terms in R equations with round brackets, just as you would do in algebra.

2. You have completed three internal assessments in one of your science papers. Each assessment was marked out of a maximum score of 20. You earned 18.5, 16 and 19 marks on the assessments. Convert your raw scores to percentages by creating a vector to hold your earned marks and then applying a single vector operation.

3. What is your average percentage mark across all three internal assessments?

4. You probably don't need to display your average mark to more than 2 decimal places. Use Google or an R textbook to find the function that will let you round a value to a given number of decimal places. Use it to generate your mean percentage mark to two decimal places.

5. Create a vector of at least six animal names. Write R commands to determine the number of characters in the longest animal name. (Read that last sentence again to make sure you understand the task requirements.) You will need to use the function `c()` to create the vector, and two other functions that we discussed in this module. 

6. To select a single element from a vector, you write the vector name followed by square brackets **[]** (the brackets should touch the name). Place the **position** of the element of interest inside the square brackets. For example:
  
```{r indexing}
animals <- c("Armadillo", "Buffalo", "Capybara", "Dingo", "Eel", "Flamingo")
animals[3]
```

In programming, an element's position in a vector is called its **index**, and the square brackets are the **index operator**. To select multiple elements, you first create a vector of the indices you want, and then pass *that vector* into the index operator. For example, to select the first and second animals in our animal vector:

```{r multiple indices}
indices <- c(1,2)
animals[indices]
```

Write an R command to display the first, third and last elements of the animal vector.


### Optional Challenge Exercise
Imagine that your animals vector had many hundreds of elements -- so many elements that you couldn't tell, just by looking, how many there were. Write R commands to select the first and last elements of a vector when you don't know the index of the last element. Test your solution on your animals vector. Hint: R has a function that will tell you how many elements are in a vector.

\

\

# Our Solutions

Click Code/Hide to toggle the answers being shown or hidden for each exercise.

\

## Exercise 1
```{r f to c, class.source = 'fold-hide'}
celsius <- (-17 - 32) * (5/9)
celsius
```

---

## Exercise 2
```{r marks, class.source = 'fold-hide'}
marks <- c(18.5, 16, 19)
percentage_marks <- marks/20
percentage_marks
```

---

## Exercise 3
```{r mean, class.source = 'fold-hide'}
mean_percentage <- mean(percentage_marks)
mean_percentage
```

---

## Exercise 4
```{r round, class.source = 'fold-hide'}
round(mean_percentage, 2)
```

---

## Exercise 5
```{r chars, class.source = 'fold-hide'}
animals <- c("Armadillo", "Buffalo", "Capybara", "Dingo", "Eel", "Flamingo")

characters_in_names <- nchar(animals)

max_characters <- max(characters_in_names)

max_characters
```

---

## Exercise 6

```{r first and last hard-coded, class.source = 'fold-hide'}

indices <- c(1,3,6)
animals[indices]
```

---

## Challenge Exercise

```{r length, class.source = 'fold-hide'}
number_of_animals <- length(animals)
first_and_last_indices <- c(1, number_of_animals)
animals[first_and_last_indices]

```
