---
title: "Zoom Notes 07: Combining Data"
date: "Semester 1, 2023"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

> #### Associated Material
>
> Module: [Module 07 - Combining data](07-combine.html)
>
> Readings
>
> - [R for Data Science Chapter 13](https://r4ds.had.co.nz/relational-data.html)


```{r setup, include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/07/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)


```


## Combining Data

How to combine datasets together

- rows
- columns
- using data relationships

### Binding

- Combining rows of datasets with same column names together with `rbind`
- Combining columns of datasets together with `cbind`
  - caution on using `cbind` and maintaining data integrity
  - use joins if possible

### Joining

Adding additional columns onto data sets by using a common identifier (key) between datasets to maintain data relationships

#### Inner join

- `inner_join` from `dplyr` will join datasets on a key and at the same time filter so that only rows where there is a matching key will be kept

#### Left/Right joins

- `left_join` and `right_join` from `dplyr` will join datasets on a key and keep all rows from the specified 'side' of the join, and filling in the non-matching entries with `NA`

#### Full joins

- `full_join` from `dplyr` will join datasets on a key and keep all rows from both datasets filling in columns of non-matching entries with `NA`

\

## Excercises



1. Using `flights` and `planes` from the `nycflights13` package, how many flights were flown in April 2013 by aircraft with each engine type? e.g. Turbo-fan, Turbo-jet, etc.

\

2. How many `flights` for the carrier `AA` with a destination (`dest`) of `IAH`, and carrier `DL` with a destination of `ATL` were late? and how many early?
    a. First create a dataframe with two columns: `carrier` and `dest`, and fill in the specific data.
    b. Use inner-join with your dataframe and `flights`, specifying `by = c("carrier", "dest")` to result in a dataframe with rows matching the conditions, assign into `flights_of_interest`.
    c. Now take `flights_of_interest` and create two new datasets, one for the late arrivals (`late`) and the second for early arrivals (`early`). How many rows in each? 

\

3. Using the datasets (`early` and `late`) you created in 2, create:
    a. a box plot for each of `early` and `late` using `arr_delay` by `carrier`
    b. add a extra column `status` onto each dataset with the values "early" for `early` and "late" for `late`
    c. combine the datasets and create a box plot of status versus air_time, but also make it a facet plot using `month`, with 6 columns and 2 rows (hint look up `facet_wrap`). Are there any trends you notice?
    

### Example solutions
    
1.

```{r zn07-soln1, class.source = "fold-hide"}
library(tidyverse)
library(nycflights13)

flights %>% 
  filter(month == 3 & year == 2013 & !is.na(arr_time)) %>%
  inner_join(planes, by = "tailnum") %>% 
  group_by(engine) %>% 
  summarise(n_flights = n())
```

--- 

2. 

```{r zn07-soln2, class.source = "fold-hide"}
carrier_dest <- data.frame(carrier = c("AA", "DL"), dest = c("IAH", "ATL"))

flights_of_interest <- inner_join(flights, carrier_dest, by = c("carrier","dest"))

late <- flights_of_interest %>% filter(arr_delay > 0)
early <- flights_of_interest %>% filter(arr_delay < 0)

nrow(late)
nrow(early)
```

---


3.

```{r zn07-soln3, class.source = "fold-hide"}
ggplot(early, aes(x = carrier, y = arr_delay)) + geom_boxplot()
ggplot(late, aes(x = carrier, y = arr_delay)) + geom_boxplot()

early <- early %>% mutate(status = "early")
late <- late %>% mutate(status = "late")

rbind(early,late) %>% 
  ggplot(aes(x = status, y = air_time)) + 
  geom_boxplot() +
  facet_wrap(vars(month), nrow = 2, ncol = 6,)

```