---
title: "Zoom Notes 07: Combining Data"
date: "Semester 1, 2023"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

> #### Associated Material
>
> Module: [Module 07 - Transforming data](07-combine.html)
>
> Readings
>
> - [R for Data Science Chapter 13](https://r4ds.had.co.nz/relational-data.html)


```{r setup, include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)


```


## Combining Data

How to combine datasets together

- rows
- columns
- using data relationships

### Binding

- Combining rows of datasets with same column names together with `rbind`
- Combining columns of datasets together with `cbind`
  - caution on using `cbind` and maintaining data integrity
  - use joins if possible

### Joining

Adding additional columns onto data sets by using a common identifier (key) between datasets to maintain data relationships

#### Inner join

- `inner_join` from `dplyr` will join datasets on a key and at the same time filter so that only rows where there is a matching key will be kept

#### Left/Right joins

- `left_join` and `right_join` from `dplyr` will join datasets on a key and keep all rows from the specified 'side' of the join, and filling in the non-matching entries with `NA`

#### Full joins

- `full_join` from `dplyr` will join datasets on a key and keep all rows from both datasets filling in columns of non-matching entries with `NA`

\

## Excercises



2. Using `flights` and `planes` from the `nycflights13` package, how many flights were flown in April 2013 by aircraft with each engine type? e.g. Turbo-fan, Turbo-jet, etc.

\

3. Using the gapminder_yearly_population_total.csv data create "long" a format dataset for 
    a. New Zealand's population from 1900 to 2000
    b. a country of your choice for the same time range

\

4. Using the datasets you created in 3, create:
    a. a line plot for each
    b. combine the datasets and create a facet plot using country with 1 column and 2 rows (hint look up `facet_grid`)
    

### Example solutions
    
2.

```{r zn05-soln2, class.source = "fold-hide"}
library(tidyverse)
library(nycflights13)

flights %>% 
  filter(month == 3 & year == 2013 & !is.na(arr_time)) %>%
  inner_join(planes, by = "tailnum") %>% 
  group_by(engine) %>% 
  summarise(n_flights = n())
```

---    