---
title: "Zoom Notes 06: Tidying Data"
date: "Semester 2, 2023"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

> #### Associated Material
>
> Module: [Module 06 - Tidying data](06-tidying.html)
>
> Readings
>
> - [R for Data Science Chapter 10](https://r4ds.had.co.nz/tibbles.html)
> - [R for Data Science Chapter 11](https://r4ds.had.co.nz/data-import.html)
> - [R for Data Science Chapter 12](https://r4ds.had.co.nz/tidy-data.html)
> - [R for Data Science Chapter 14](https://r4ds.had.co.nz/strings.html)

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/06/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)


```

## Cleaning and Tidying

### Cleaning data values

- Checking data types on data import
- Creating consistent missing data
- Converting characters to numeric with `as.numeric`

### Cleaning dataframe column names

- Using the `janitor` package to create cleaned dataframe column names

### Tidying

Tidy data principles:

1. Each column is a variable or property that is being measured 
2. Each row is an observation
3. A single cell should contain a single piece of information

## Reshaping data

### Wide to long

- Use `pivot_longer` from the `tidyr` package

### Long to wide

- Use `pivot_wider` from the `tidyr` package

\

## Excercises

1. Using the rodents_untidy.csv, read the data in and then:
    a. use janitor to change the column names to the "small_camel" case
    b. clean the data "weight" column
        - convert "?" and -999 to NA
        - convert the column to numeric
    c. split the plotLocation column into two columns: "plot" and "location"
    d. make the "plot" column into a numeric
    e. **challenge**: fix the data in the species column for plot 3. The genus and species columns had accidentally been combined
    
\

2. Using the gapminder_yearly_population_total.csv data create "long" a format dataset for 
    a. New Zealand's population from 1900 to 2000
    b. a country of your choice for the same time range

### Example solutions

1.

```{r zn06-soln1, class.source = "fold-hide"}
library(tidyverse)
rodents <- read_csv("data/rodents_untidy.csv")
# a)
rodents <- rodents %>% janitor::clean_names(case = "small_camel")

# b)
rodents$weight[rodents$weight %in% c("-999", "?")] <- NA
rodents$weight <- as.numeric(rodents$weight)

# c)
rodents <- rodents %>% separate(plotLocation, into = c("plot", "location"), sep = "_")

# d)
rodents$plot <- as.numeric(rodents$plot)

# e)
plot1_2 <- rodents %>% 
  filter(plot ==1 | plot == 2)

plot_3 <- rodents %>% 
  filter(plot == 3) %>% 
  select(-genus) %>% 
  separate(species, into = c("genus", "species"), sep=" ")

rodents_clean <- rbind(plot1_2, plot_3)
rodents_clean
```

---



2.

```{r zn06-soln2, class.source = "fold-hide", eval = FALSE}
library(tidyverse)
gapminder_yearly_pop <- read_csv("data/gapminder_yearly_population_millions_total.csv") %>% mutate(across(-country, ))

gapminder_yearly_pop_long <- gapminder_yearly_pop %>% pivot_longer(-country, names_to = "year", values_to = "population_millions") %>% mutate(year = as.numeric(year))

# a)
nz_1900_2000 <- gapminder_yearly_pop_long %>% 
  filter(between(year, 1900, 2000)) %>% 
  filter(country == "New Zealand")

# b)
jamaica_1900_2000 <- gapminder_yearly_pop_long %>% 
  filter(between(year, 1900, 2000)) %>% 
  filter(country == "Jamaica")
```

---



