---
title: "Zoom Notes 05: Transforming Data"
date: "Semester 2, 2022"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

> #### Associated Material
>
> Module: [Module 05 - Transforming data](05-transform.html)
>
> Readings
>
> - [R for Data Science Chapter 10](https://r4ds.had.co.nz/tibbles.html)
> - [R for Data Science Chapter 11](https://r4ds.had.co.nz/data-import.html)
> - [R for Data Science Chapter 12](https://r4ds.had.co.nz/tidy-data.html)
> - [R for Data Science Chapter 13](https://r4ds.had.co.nz/relational-data.html)
> - [R for Data Science Chapter 14](https://r4ds.had.co.nz/strings.html)

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)


```

## Cleaning and Tidying

### Cleaning data values

- Checking data types on data import
- Creating consistent missing data
- Converting characters to numeric with `as.numeric`

### Cleaning dataframe column names

- Using the `janitor` package to create cleaned dataframe column names

### Tidying

Tidy data principles:

1. Each column is a variable or property that is being measured 
2. Each row is an observation
3. A single cell should contain a single piece of information

## Reshaping data

### Wide to long

- Use `pivot_longer` from the `tidyr` package

### Long to wide

- Use `pivot_wider` from the `tidyr` package

\

## Combining Data

How to combine datasets together

- rows
- columns
- using data relationships

### Binding

- Combining rows of datasets with same column names together with `rbind`
- Combining columns of datasets together with `cbind`
  - caution on using `cbind` and maintaining data integrity
  - use joins if possible

### Joining

Adding additional columns onto data sets by using a common identifier (key) between datasets to maintain data relationships

#### Inner join

- `inner_join` from `dplyr` will join datasets on a key and at the same time filter so that only rows where there is a matching key will be kept

#### Left/Right joins

- `left_join` and `right_join` from `dplyr` will join datasets on a key and keep all rows from the specified 'side' of the join, and filling in the non-matching entries with `NA`

#### Full joins

- `full_join` from `dplyr` will join datasets on a key and keep all rows from both datasets filling in columns of non-matching entries with `NA`

\

## Excercises

1. Using the rodents_untidy.csv, read the data in and then:
    a. use janitor to change the column names to the "small_camel" case
    b. clean the data "weight" column
        - convert "?" and -999 to NA
        - convert the column to numeric
    c. split the plotLocation column into two columns: "plot" and "location"
    d. make the "plot" column into a numeric
    e. **challenge**: fix the data in the species column for plot 3. The genus and species columns had accidentally been combined
    
\

2. Using `flights` and `planes` from the `nycflights13` package, how many flights were flown in April 2013 by aircraft with each engine type? e.g. Turbo-fan, Turbo-jet, etc.

\

3. Using the gapminder_yearly_population_total.csv data create "long" a format dataset for 
    a. New Zealand's population from 1900 to 2000
    b. a country of your choice for the same time range

\

4. Using the datasets you created in 3, create:
    a. a line plot for each
    b. combine the datasets and create a facet plot using country with 1 column and 2 rows (hint look up `facet_grid`)

### Example solutions

1.

```{r zn05-soln1, class.source = "fold-hide"}
library(tidyverse)
rodents <- read_csv("data/rodents_untidy.csv")
# a)
rodents <- rodents %>% janitor::clean_names(case = "small_camel")

# b)
rodents$weight[rodents$weight %in% c("-999", "?")] <- NA
rodents$weight <- as.numeric(rodents$weight)

# c)
rodents <- rodents %>% separate(plotLocation, into = c("plot", "location"), sep = "_")

# d)
rodents$plot <- as.numeric(rodents$plot)

# e)
plot1_2 <- rodents %>% 
  filter(plot ==1 | plot == 2)

plot_3 <- rodents %>% 
  filter(plot == 3) %>% 
  select(-genus) %>% 
  separate(species, into = c("genus", "species"), sep=" ")

rodents_clean <- rbind(plot1_2, plot_3)
rodents_clean
```

---

2.

```{r zn05-soln2, class.source = "fold-hide"}
library(tidyverse)
library(nycflights13)

flights %>% 
  filter(month == 3 & year == 2013 & !is.na(arr_time)) %>%
  inner_join(planes, by = "tailnum") %>% 
  group_by(engine) %>% 
  summarise(n_flights = n())
```

---

3.

```{r zn05-soln3, class.source = "fold-hide", eval = FALSE}
library(tidyverse)
gapminder_yearly_pop <- read_csv("data/gapminder_yearly_population_millions_total.csv") %>% mutate(across(-country, ))

gapminder_yearly_pop_long <- gapminder_yearly_pop %>% pivot_longer(-country, names_to = "year", values_to = "population_millions") %>% mutate(year = as.numeric(year))

# a)
nz_1900_2000 <- gapminder_yearly_pop_long %>% 
  filter(between(year, 1900, 2000)) %>% 
  filter(country == "New Zealand")

# b)
jamaica_1900_2000 <- gapminder_yearly_pop_long %>% 
  filter(between(year, 1900, 2000)) %>% 
  filter(country == "Jamaica")
```

---

4.

```{r zn05-soln4, class.source = "fold-hide", eval = FALSE}
# a)
nz_1900_2000 %>% ggplot(aes(x = year, y = population_millions)) + geom_line() + labs(title = "NZ population 1900-2000")

jamaica_1900_2000 %>% ggplot(aes(x = year, y = population_millions)) + geom_line() + labs(title = "Jamaica population 1900-2000")

# b)
rbind(nz_1900_2000, jamaica_1900_2000) %>% 
  ggplot(aes(x = year, y = population_millions)) + 
  geom_line() + 
  facet_grid(rows = "country") + 
  labs(title = " NZ vs Jamaica population 1900-2000")
```

