---
title: "Zoom Notes 10: Workflows"
date: "Semester 1, 2023"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/09/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)
```

> #### Associated Material
>
> Module: [Module 10 - Workflows](10-workflows.html)

## Getting organised

- Create a structure
- Create a README
- Choose a license

The `usethis` package can help you, and provides many helper function for creating and setting up projects and packages

```{r, eval = FALSE}
# Create and open a new package
usethis::create_project(path = "path/to/new/project")

# Create a markdown README
usethis::use_readme_md()

# Create an RMarkdown README
usethis::use_readme_rmd()

# Create a new R script in your project in directory R/
usethis::use_r(name = "myfile.R")

```

### Configuration

It's worth taking some time to configure options in your environment to make life easier for yourself.

The `.Rprofile` file is used to store these configurations. You can open it with `usethis::edit_r_profile()`.

- Don't load 'analysis' dependant packages in your Rprofile

Learn some of the keyboard shortcuts in RStudio to improve your productivity. <kbd>Alt</kbd> + <kbd>Shift</kbd> + <kbd>K</kbd> is the keyboard shortcut to show you the keyboard shortcuts.

### Version control

Using an automated version control system such as `git` is highly recommended. It can help you manage different versions of your code and show differences or let you jump between versions.

## Debugging

General process:

1. Do I recognise the message?
2. Re-look at the command I ran looking for
    - typos
    - missing syntax (e.g. brackets, semicolons, etc.)
    - correct naming of things
3. Is the input for my command what I expect it is?
4. Did the previous command run properly?
    - if not jump to 1. for the previous command


## Testing

Test your code as you go. It can be useful to adopt an automated testing solution - such as the `testthat` package to create tests that can be automatically run.
