---
title: "Visualising"
date: "Semester 1, 2022"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---
  
```{r setup, include=FALSE}
library(ggplot2)

library(knitr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)
```

> #### Data Focus
  >
  > Modules and Handouts 2--6 cover topics that are part of the data analysis journey and are all interrelated. Each module will introduce new content and expand on the material covered in previous modules. For example, in Module 2 we introduce the basics of the R plotting systems. Subsequent handouts include more advanced plotting options and techniques.

\

\


Zoom Notes: [Zoom Notes 02 - Visualising Data](zoom_notes_02.html)

Associated Readings:
  
- [R for Data Science - Chapter 12](https://r4ds.had.co.nz/tidy-data.html)
- [R for Data Science - Chapter 11](https://r4ds.had.co.nz/data-import.html)
- [R for Data Science - Chapter 3](https://r4ds.had.co.nz/data-visualisation.html)


# Topics

## Tabular Data

- Edit in Excel
- Export to plain csv (not UTF-8)
- Column headers in first row for every column
- One row for each data record

## Folder setup

- Eventually, use Projects to allow RStudio to manage metadata
- Separate folders for data files, images, scripts, Rmd, etc.

## Importing a data file

- read.csv
- Set stringsAsFactors = TRUE to import categorical variables correctly
- After we meet the tidyverse, can also use read_csv

## Checking your imported data

- head
- tail
- str
- Confirm that each column is of the correct type


## Selecting and using columns of data

- Select an individual column with the $ operator


## Base R plots

- hist for frequency distributions
- boxplots show central tendency and variability
- formulas have the form *dependent variable* **~** *linear model of independent variables*


## Plotting with ggplot2

- Use library `ggplot` contained in package ggplot2
- Complex syntax based on Grammar of Graphics (from computer science)
- Plots built in layers


## Building a graph

- All plots begin with call to ggplot, passing in a data frame
- Mappings define relationships between elements of the data and visual features on the plot
- Use function `aes` to define a mapping
- Assign column names to `aes` arguments `x` and `y` to define graph axes
- Many available arguments to `aes`; part of the ggplot syntax
- Select a **geometry** to determine the kind of plot (e.g. bar graph, scatterplot, line graph, etc.)
- Additional layers define axes labels, title, legends, and fonts
- Combine ggplot layer sub-commands with `+`

# Practice Exercises

To practice what we have learned in Module 02, we will use "Palmer's Penguins", a real data set from the Palmer Station Long Term Ecological Research program (https://allisonhorst.github.io/palmerpenguins/articles/intro.html). These data are size measurements for three penguin species -- Chinstrap, Gentoo and Adelie -- on three islands in Antarctica.

You will need to install the package that contains the data frame (code shown below). Then work through each of the exercises. If you have any questions, email us or send us a message in MS Teams.

## Get the Penguin Data

```{r install, eval=FALSE}

# Install the package (do once on any computer)
install.packages("palmerpenguins")
```

```{r load}
# Load the library (do at the start of every RStudio session)
library(palmerpenguins)

# Check the data - the data frame name is penguins
str(penguins)
```

The output from `str(penguins`) indicates that three of the columns in the data frame are **Factors**. In R, a factor is a categorical variable, usually corresponding to an experimental factor. Although factors look like strings, a factor is restricted to a specific set of legal values, which R infers when the data are loaded. The legal values are called **levels**, and correspond to the different groups or conditions represented by the factor. For example, column `penguins$sex` is a factor with levels "female" and "male".

## Exercises
1. When our data sets have factors, we often use functions `levels` and `table`. Use Google or your favourite text book to explore these functions. Use them to solve the next two exercises.

2. What are the three different levels of the **species** factor? What are the three different levels of the **island** factor?

```{r levels, echo=FALSE, eval=FALSE}
levels(penguins$species)
levels(penguins$island)
```


3. How many observations are there in the data frame for each of the three species? How many observations are there in the data frame for each of the three islands?

```{r table, echo=FALSE, eval=FALSE}
table(penguins$species)
table(penguins$island)
```

4. This code computes the mean and standard deviation of bill length collapsed across islands, species, and sex? (To collapse across factors means to ignore the fact that there are different levels; just bung all the data together.) We add the argument `na.rm = TRUE` to tell R to ignore any NA (missing) values.

```{r mean and sd demo, eval = FALSE}
mean(penguins$bill_length_mm, na.rm = TRUE)
sd(penguins$bill_length_mm, na.rm = TRUE)
```

What are the mean and standard deviation of flipper length, collapsed across islands, species, and sex? 

```{r mean and sd, echo=FALSE, eval=FALSE}
mean(penguins$flipper_length_mm, na.rm = TRUE)
sd(penguins$flipper_length_mm, na.rm = TRUE)
```

5. On average, who has longer flippers, male penguins or female penguins?
```{r make vs. female, echo=FALSE, eval=FALSE}
males <- subset(penguins, sex == "male")
females <- subset(penguins, sex == "female")
mean(males$flipper_length_mm)
mean(females$flipper_length_mm)
```


6. Using base R, generate a histogram showing the distribution of body mass, collapsed across island, species and sex. How would you describe the distribution?

```{r hist, echo=FALSE, eval=FALSE}
hist(penguins$body_mass_g)
```

7. Using ggplot, generate a scatterplot illustrating the relationship between bill length and body mass, collapsed across species, island and sex. Remember to load the library with `library(ggplot2)` before first use.How would you describe the pattern?

```{r scatterplot, echo=FALSE, include=FALSE}

library(ggplot2)

ggplot(data = penguins) +
  geom_point(mapping = aes(x = bill_length_mm, y = body_mass_g))

```


8. Modify your plot from Exercise 7 so that penguins from the different islands are drawn in different colours. Which island seems to have the heaviest penguins? Without looking any further at the data, formulate at least two possible explanations for the pattern.

```{r color, echo=FALSE, eval=FALSE}
ggplot(data = penguins) +
  geom_point(mapping = aes(x = bill_length_mm, y = body_mass_g, color = island))

# Possible explanations: 1) More food on that island 2) Bigger species of penguin on that island

```

9. Using facet_wrap, generate a figure that shows the bill length by body mass scatterplot separately for each species.

```{r facet wrap, echo=FALSE, eval=FALSE}
ggplot(data = penguins) +
  geom_point(mapping = aes(x = bill_length_mm, y = body_mass_g)) +
  facet_wrap(~ species, nrow = 2)
```

10. Using geom_smooth, illustrate the bill length by body mass relationship as a smoothed curve.

```{r geom_smooth, echo=FALSE, eval=FALSE}
ggplot(data = penguins) +
  geom_smooth(mapping = aes(x = bill_length_mm, y = body_mass_g))
```

11. Using ggplot, generate a boxplot comparing body mass for the three different species of penguin, and having each of the three boxes drawn in a different colour. What information is missing from this figure that was provided Exercise 8? What information is easier to see in this figure than in Exercise 8?

```{r ggplot boxplot, echo=FALSE, eval=FALSE}
ggplot(data = penguins) +
  geom_boxplot(mapping = aes(x = island , y = body_mass_g, color = island))

```

12. Using ggplot, duplicate this figure:

```{r barplot with fill and dodge, echo=FALSE}
ggplot(data = penguins) +
  geom_bar(mapping = aes(x = island, fill=species), position="dodge")
```


