---
title: "Zoom Notes 04: Communicating Data"
date: "Semester 1, 2023"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
    code_folding: show
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.path = "figures/04/", # use only for single Rmd files
  collapse = TRUE,
  echo = TRUE
)


```

> #### Associated Material
>
> Module: [Module 04 - Communicate](04-communicate.html)
>
> Readings:
>
> -   [R for Data Science - Chapter 27](https://r4ds.had.co.nz/r-markdown.html)
> -   [R for Data Science - Chapter 29](https://r4ds.had.co.nz/r-markdown-formats.html)
> -   [Reference: RMarkdown - the definitive guide](https://bookdown.org/yihui/rmarkdown/)
> -   [RMarkdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

## What is RMarkdown

There are three main components to an RMarkdown document

1.  The **YAML** header which is surrounded by `---`s and provides information for the compiling process
2.  R code **chunks** which are surrounded by ```` ``` ````s
3.  Text which can be formatted using the Markdown language.


A reference guide of RMarkdown syntax can be found through `Help` -> `Cheat Sheets` -> `R Markdown Reference Guide` in the RStudio menu.

### Example Rmd

Create your own RMarkdown document now from the template. To do this go `File` -\> `New File` -\> `R Markdown`.

### Knitting

"Knit" or compile the document using the `knit` button or <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>K</kbd>.

\

## Markdown syntax

Refer to the [RMarkdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

\

## Code Chunks

Code chunks look like:

```{r echo = FALSE, comment = ""}
cat(htmltools::includeText("scripts/code_chunk.Rmd"))
```

- Code chunks are executed as part of the document "knitting" and the code and output embedded in the document

- Be aware of the working directory (don't use `setwd()`). Ideally use the `here` package to manage paths.

\

## Images, Figures, and Tables

- Images can be included though markdown, or by using the `include_graphics()` from `knitr` in a code chunk - the latter provides more options for customising for size/position.

- Figures are created from code chunks and chunk options are used to control size/position

- Tables can be manually created using the markdown syntax or created automatically from your data using the `kable()` from `knitr`.

Further customising of `kable` tables can be done with the `kableExtra` package.

\

## Citations

Citations can be inserted into an RMarkdown document. [This document from RStudio](https://rstudio.github.io/visual-markdown-editing/citations.html) goes through how to do it


## Exercises

1. Create a new RMarkdown file called with the output being HTML.
    a. fill in your name and the title `My Report`
2. Save the file as `my_report.Rmd` and knit it.
3. Modify the `setup` code chunk to load the `tidyverse` package
4. Remove the template content starting on the line with `## R Markdown` and below.
5. Create a code chunk to read in the gapminder data used in [Module 2 - Visualising](02-visualise.html)
    a. _Watch out for your file paths relative to your Rmd file _
6. Choose two countries and filter the data to only include them
7. Make a plot of life expectancy versus gdp per capita
8. In the text areas of the RMarkdown note some observations about your plot. Include:
    a. a level 2 heading
    b. something in italics
    c. something in bold
9. Underneath your plot include a table of the data for your countries
